---
title: "HST.953x Workshop 2.09: IAC Case Study"
author: "H. David Shea"
date: "`r format(Sys.time(), '%d %b %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    fig.align: center
    fig_caption: yes
    fig_height: 5
    fig_width: 9
    highlight: tango
    theme: united
    #toc: yes
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)
library(knitr)
library(MIMICbook)
library(Hmisc)

knitr::opts_chunk$set(
    connection = "mimic3", # automatically uses this connection in sql chunks 
    comment = "#>", 
    collapse = TRUE, 
    message = FALSE,
    fig.width = 8,
    fig.asp = ((1 + sqrt(5)) / 2) - 1, # the golden ratio - technically, the b proportion of a+b when a is 1
    out.width = "70%",
    fig.align = "center"
)

base_dir <- here::here("")
```

## Description

This document contains the write up and code for my version of interpretation of the data from the study of whether indwelling arterial catheters (IAC) have any effect on patient mortality. The study objective was:

“To estimate the effect that administration of IAC during an ICU admission has on 28 day mortality in patients within the MIMIC II study who received mechanical ventilation, while adjusting for age, gender, severity of illness and comorbidities.”

## Base data set

Import dataset from **PhysioNet** - raw data file downloaded in previous exercise.

```{r basedata, message=FALSE, warning=FALSE}
fnm <- fs::path(base_dir, "course_exercises/exploratory_data_analysis/aline_full_cohort_data.csv")
dat <- read.csv(fnm)
rm(fnm)

is01_factor_column <- function(x) {
  v <- unique(x[!is.na(x)])
  ((length(v) == 1) & (v[1] %in% c(0,1))) | ((length(v) == 2) & (v[1] %in% c(0,1)) & (v[2] %in% c(0,1)))
}

# convert to factors where appropriate
dat <- dat %>% 
  mutate(across(where(is01_factor_column), as.factor))
dat$service_unit <- as.factor(dat$service_unit)
dat$day_icu_intime <- as.factor(dat$day_icu_intime)
```

## Exploratory Data Analysis

Here we examine:

* full cohort characteristics
* chracteristics of IAC versus non-IAC parts of the cohort
* chracteristics of portion of the cohort who were alive after 28 days versus those who died

```{r eda, message=FALSE, warning=FALSE}
tab_full <- produce.table1(dat)
kable(tab_full, caption = "Full cohort characteristics")

dat_by_aline <- split(dat, dat$aline_flg)
labs <- attr(produce.table1(dat_by_aline[[1]]), "labels")
tab_by_aline <- cbind(
  produce.table1(dat_by_aline[[1]], labels = labs),
  produce.table1(dat_by_aline[[2]], labels = labs)
)
colnames(tab_by_aline) <- paste0("Average (SD), or N (%), ", c("No IAC", "IAC"))
kable(tab_by_aline, caption = "Patient characteristics stratified by IAC administration")

dat_by_28day <- split(dat, dat$day_28_flg)
labs <- attr(produce.table1(dat_by_28day[[1]]), "labels")
tab_by_28day <- cbind(
  produce.table1(dat_by_28day[[1]], labels = labs),
  produce.table1(dat_by_28day[[2]], labels = labs)
)
colnames(tab_by_28day) <- paste0("Average (SD), or N (%), ", c("Alive", "Dead"))
kable(tab_by_28day, caption = "Patient characteristics stratified by 28 day mortality")
```

## Basic Mortality Analysis

In the tables above, we see that 170 of the 984 (17.2 %) subjects receiving IAC died within 28 days, whereas 113 of 792 (14.2 %) died within 28 days in the no-IAC group. In a univariate analysis, we can assess if the lower rate of mortality is statistically significant by fitting a single covariate `aline_flg` logistic regression.

Note:  All of the variables that we will use in the following analyses are binary or categorical except for age and the disease severity indicators (`sapsi_first` and `sofa_first`) which are continuous.  While, our previous work does show that age has a linear relation to 28 day mortality, it has also shown that SAPS and SOFA scores do not.  These will be converted to categorical variables for the analysis.

```{r logreg_mort, message=FALSE, warning=FALSE}
dat$sapsi_cat <- cut2(dat$sapsi_first, g = 5)
dat$sofa_cat <- cut2(dat$sofa_first, g = 5)

mort.glm <- glm(day_28_flg ~ aline_flg, data = dat, family = "binomial")

exp(mort.glm$coef[-1])

exp(confint(mort.glm)[-1,])

summary(mort.glm)$coef
```

The part of the cohort who received IAC had over a 25% increase in odds of 28 day mortality when compared to those who did not receive IAC. However, the confidence interval includes one (`ln(1) == 0`), so we would expect the p-value would be >0.05 which indeed shows in the coefficient summary for `aline_flg` at about 0.09.

As seen in the exploratory analyses, there are likely several important covariates that differed among those who received IAC and those who did not which may serve as confounders which could impact the view from this univariate analysis depending on the situation.

We will employ backwards step-wise elimination to get a better feel for the potential impact of confounding variables.

## Step-wise Elimination

The variables under consideration for a full model are `aline_flg`, `age`, `gender_num`, disease severity indications (`sapsi_first`, `sofa_first`), ICU service type (`service_unit`), and comorbidities (`chf_flg`, `afib_flg`, `renal_flg`, `liver_flg`, `copd_flg`, `cad_flg`, `stroke_flg`, `mal_flg` and `resp_flg`).

```{r full_model_glm, message=FALSE, warning=FALSE}
mort.full.glm <- glm(day_28_flg ~ aline_flg + age + gender_num + sapsi_cat + 
                       sofa_cat + service_unit + chf_flg + afib_flg + renal_flg + 
                       liver_flg + copd_flg + cad_flg + stroke_flg + mal_flg + resp_flg,
                     data = dat, family = "binomial")

summary(mort.full.glm)
```

Now we use `drop1` with a Chi-square test to determine the variable with the least significance (i.e., highest Pr(>Chi) _p_-value) .

```{r full_model_drop1, message=FALSE, warning=FALSE}
drop1(mort.full.glm, test = "Chisq")
```

While `aline_flg` has the largest _p_-value, this variable is our primary discriminator for the study, so we drop the variable with the next highest _p_-value `cad_flg` the coronary artery disease indicator variable.

```{r d1_model_glm, message=FALSE, warning=FALSE}
mort.d1.glm <- update(mort.full.glm, .~. - cad_flg)

drop1(mort.d1.glm, test = "Chisq")
```

The following are the remaining iterations until only _p_-values < 0.05 exist.

`chf_flg` the congestive heart failure indicator variable.

```{r d2_model_glm, message=FALSE, warning=FALSE}
mort.d2.glm <- update(mort.d1.glm, .~. - chf_flg)

drop1(mort.d2.glm, test = "Chisq")
```

`gender_num` the patient gender indicator variable (1 = male; 0 = female).

```{r d3_model_glm, message=FALSE, warning=FALSE}
mort.d3.glm <- update(mort.d2.glm, .~. - gender_num)

drop1(mort.d3.glm, test = "Chisq")
```

`opd_flg` the chronic obstructive pulmonary disease indicator variable.

```{r d4_model_glm, message=FALSE, warning=FALSE}
mort.d4.glm <- update(mort.d3.glm, .~. - copd_flg)

drop1(mort.d4.glm, test = "Chisq")
```

`liver_flg` the liver disease indicator variable.

```{r d5_model_glm, message=FALSE, warning=FALSE}
mort.d5.glm <- update(mort.d4.glm, .~. - liver_flg)

drop1(mort.d5.glm, test = "Chisq")
```

`sofa_cat` the SOFA score category variable.

```{r d6_model_glm, message=FALSE, warning=FALSE}
mort.d6.glm <- update(mort.d5.glm, .~. - sofa_cat)

drop1(mort.d6.glm, test = "Chisq")
```

`renal_flg` the chronic renal disease indicator variable.

```{r d7_model_glm, message=FALSE, warning=FALSE}
mort.d7.glm <- update(mort.d6.glm, .~. - renal_flg)

drop1(mort.d7.glm, test = "Chisq")
```

`service_unit` the type of ICU service unit indicator variable.

```{r d8_model_glm, message=FALSE, warning=FALSE}
mort.d8.glm <- update(mort.d7.glm, .~. - service_unit)

drop1(mort.d8.glm, test = "Chisq")
```

Final Model

```{r final_model_glm, message=FALSE, warning=FALSE}
mort.final.glm <- mort.d8.glm

summary(mort.final.glm)

mort_final_tab <- cbind(round(exp(mort.final.glm$coef[-1]), 2),
                        round(exp(confint(mort.final.glm)[-1,]), 2),
                        round(summary(mort.final.glm)$coefficients[-1, 4], 3))
rownames(mort_final_tab) <-
  c(
    "IAC",
    "Age (per year increase)",
    "SAPSI [12–14)* (relative to SAPSI <12)",
    "SAPSI [14–16)*",
    "SAPSI [16–19)*",
    "SAPSI [19–32]*",
    "Atrial fibrillation",
    "Stroke",
    "Malignancy",
    "Non-COPD respiratory disease"
  )
colnames(mort_final_tab) <-
  c("AOR", "Lower 95% CI", "Upper 95% CI", "p-value")
kable(mort_final_tab, 
      caption = "Multivariable logistic regression analysis for mortality at 28 days outcome (final model)")
```
