---
title: "Length of stay in the ICU"
author: "H. David Shea"
date: "`r format(Sys.time(), '%d %b %Y')`"
output:
  pdf_document:
    toc: yes
---

(Note:  Updated and modified from the hst953-edx version authored by Tom Pollard.)

```{r setup, include = FALSE}
library(DBI)
library(RSQLite)
library(tidyverse)
library(lubridate)
library(patchwork)

knitr::opts_chunk$set(
    connection = "mimic3", # automatically uses this connection in sql chunks 
    comment = "#>", 
    collapse = TRUE, 
    message = FALSE,
    fig.width = 8,
    fig.asp = ((1 + sqrt(5)) / 2) - 1, # the golden ratio - technically, the b proportion of a+b when a is 1
    out.width = "70%",
    fig.align = "center"
)
```
 
## Source MIMIC-III data

In the original version (on the hst953-edx github site), they used the MIMIC-III demo version directly loaded.  Now, I have in *mimic_base_dir*/database/mimic3.db the SQLite version of the full MIMIC-III v1.4 database loaded.  I'll use that in the processing below - with some pre-coded inclusion criteria to extract just the demo data.  The following code chunk attaches the database and loads auxiliary functions (available in *mimic_base_dir*/mimic_concepts) for extracting database data (`db_functions.R`) and for doing some MIMIC data interpretation and pre-processing (`mimic3_meta_data.R`).

```{r source_aux_functions, include=TRUE}
base_dir <- here::here("")
mimic_base_dir <- fs::path(base_dir,"../MIMIC-research")
db_file <- fs::path(mimic_base_dir, "database/mimic3.db")
if(dbCanConnect(RSQLite::SQLite(), db_file)) {
    mimic3 <- dbConnect(RSQLite::SQLite(), db_file)
} else {
    stop(str_c("Database file: ", db_file, " not found.", sep=""))
}
source(fs::path(mimic_base_dir, "mimic_concepts/db_functions.R"))
source(fs::path(mimic_base_dir, "mimic_concepts/mimic3_meta_data.R"))
```

## Extract ICU stay data

```{r load_icu_data, include=TRUE}
data <- db_get_icustays(mimic3) %>% select(SUBJECT_ID, HADM_ID, LOS)
 
str(data)
```

## Calculate median stay length

This document shows how RMarkdown can be used to create a reproducible analysis using MIMIC-III (version 1.4). Let's calculate the median length of stay in the ICU and then include this value in our document.

```{r calculate_mean_los, include=FALSE}
avg_los <- median(data$LOS, na.rm=TRUE)
rounded_avg_los <-round(avg_los, digits = 2)
```

So the median length of stay in the ICU is `r avg_los` days. Rounded to two decimal places, this is `r rounded_avg_los` days.

## Plot the data

We can plot the distribution of length of stay using `ggplot` and `geom_histogram`:  

```{r plot_los, echo=FALSE, include=TRUE,  warning = FALSE}
data %>%
   ggplot(aes(x = LOS)) +
   geom_histogram(
      binwidth = 1,
      fill = "#FF9999",
      color = "white",
      na.rm = TRUE) +
   scale_x_continuous(limits = c(0, 25), breaks = seq(0, 25, by = 5)) +
   labs(x = "Length of stay in the ICU, days.", y = NULL)
```
